---
title: "Kalman Filter for State Space Models"
author: "Alex Hubbard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kalman Filter for State Space Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE, include = TRUE, message = FALSE, warning = FALSE, eval = FALSE
)
```

`kalmanfilter` is an `Rcpp` implementation of the multivariate Kalman filter for state space models that can handle missing values and exogenous data in the observation and state equations. The Kalman filter is a method to compute the optimal estimator of the unobserved state vector in a time series.

# The State Space Model

The state space model is written as
$$
Y_t = A + H \beta + B^o X^o_t + e_t, e_t \sim N(0, R)\\
\beta_t = D + F \beta_{t-1} + B^s X^s_t + u_t, u_t \sim N(0, Q)
$$
where the first equation is the observation equation and the second equation is the state equation. $A$ is the observation intercept matrix, $H$ is the observation matrix, $X^o_t$ is optional exogenous data in the observation equation, $e_t \sim N(0, R)$ is the observation error, and $R$ is the observation error-covariance matrix. $\beta_t$ is the vector of the state components, $D$ is the state intercept matrix, $F$ is the transition matrix, $X^s_t$ is optional exogenous data in the state equation, $v_t \sim N(0, Q_{s_t})$ is the state error, and $Q$ is the state error-covariance matrix. 

# The Kalman Filter

The Kalman filter is a forward two-stage procedure that is the optimal linear filter based on the multivariate normal distribution. The "forwardness" of the filter means that it uses only data up to time $t$ to make an inference on the unobserved components at time $t$ and does peak into the future to make that inference.

## Prediction Stage

The first stage is the prediction stage, which makes predictions of the state components based on information up to time $t-1$. This stage is made up of four equations, the first is the prediction of the state components based on its time series properties

$$
B_{t|t-1} = D + F \beta_{t-1|t-1} + B^s X^s_t
$$

Second, is the prediction of the covariance matrix of the state components

$$
P_{t|t-1} = F P_{t_1|t-1} F^{\prime} + Q
$$

Third, is the prediction error of the time series of interest

$$
\eta_{t|t-1} = Y_t - (A + H \beta_{t|t-1} + B^o X^o_t)
$$

And finally, we have the variance of the prediction error

$$
f_{t|t-1} = H P_{t|t-1} H^{\prime} + R
$$

## Updating Stage

The second stage is the updating stage, which makes predictions based on all information up to time $t$. It consists of three equations. The first equation is the prediction of the state components based on the the full information set

$$
\beta_{t|t} = B_{t|t-1} + K_t \eta_{t|t-1}
$$
where $K_t$ is the Kalman gain, which determines the optimal weight to give new information in making predictions about $\beta_t$ and is the second equation

$$
K_t = P_{t|t-1} H^{\prime} f_{t|t-1}^{-1}
$$
The last equation is the updating of the covariance matrix of the state components

$$
P_{t|t} = P_{t|t-1} - K_t H^{\prime} P_{t|t-1}
$$
The seven equations above, make up the full Kalman filter routine. If $Y_t$ is missing for any observation, then 

$$
B_{t|t} = B_{t|t-1} \\
P_{t|t} = P_{t|t-1} \\
K_t = 0 \\
f_{t|t-1} = \infty
$$

## Kalman Smoothing

Once the Kalman filter is applied to the data, a smoothing procedure can be applied in the backward direction to make a better inference of the state components based on the entire data set. Unlike the filter, the smoother does peak into the future to make an inference of the state components at time $t$. This procedure consists of only two equations. 

The first equation updates the prediction of the state components based on all the available information

$$
\beta_{t|T} = \beta_{t|t} + P_{t|t} F^{\prime} P_{t|t}^{-1} (\beta_{t+1|T} - \beta_{t+1|t})
$$

The second equation updates the covariance matrix of the state components based on all the available information

$$
P_{t|T} = P_{t|t} + P_{t|t} F^{\prime} P_{t+1|t}^{-1} (P_{t+1|T} - P_{t+1|t}) P_{t+1|t}^{-1 \prime} F P_{t|t}^{\prime}
$$

# Estimation

To estimate a model with the Kalman filter via maximum likelihood, the log likelihood is given by

$$
ln(L(\theta)) = -\frac{1}{2} \sum_{t=1}^T \ln(|f_{t|t-1}|) - \frac{1}{2}\sum_{t=1}^T \eta_{t|t-1}^{\prime} f_{t|t-1}^{-1} \eta_{t|t-1} 
$$

# Implementation 

To use the package, use the function `kalman_filter` as 
```{r}
kf = kalman_filter(ssm, yt, Xo, Xs, w, smooth)
```

where `ssm` is a list object with matrices that describes the state space model. `ssm` should contain matrices named `B0` for the initial guess of the unobserved components, `P0` for the initial guess of the unobserved components covariance matrix, `Dm` for the constant in the state equation, `Am` for the constant in the observation equation, `Fm` for the state equation transition matrix, `Hm` for the observation matrix in the observation equation, `Qm` for the covariance matrix of the errors in the state equation, `Rm` for the covariance matrix of the errors in the observation equation, `betaO` for the coefficients on the exogenous data in the observation matrix, and `betaS` for the coefficients on the exogenous data in the state matrix. `betaO` and `betaS` are optional.

`yt` is an $N_y x T$ matrix, where $N_y$ is the number of variables and $T$ is the number of time periods. 

`Xo` is an optional $N_o x T$ matrix of optional exogenous data in the observation equation, where $N_o$ is the number of exogenous observation variables.

`Xs` is an optional $N_s x T$ matrix of optional exogenous data in the state equation, where $N_s$ is the number of exogenous state variables.

`w` is an optional $T x 1$ matrix of optional weights applied to the likelihood.

`ssm[["B0"]]`, `ssm[["P0"]]`, `ssm[["Qm"]]` are $N_{\beta} x N_{\beta}$ matrix, where $N_{\beta}$ is the number of unobserved variables in the state equation.

`ssm[["Dm"]]` is an $N_{\beta} x 1$ matrix.

`ssm[["Am"]]` is an $N_y x 1$ matrix.

`ssm[["Fm"]]` is an $N_{\beta} x P$ matrix, where $P$ is the number of lags in the state equation.

`ssm[["Hm"]]` is an $N_y x N_{\beta}$ matrix.

`ssm[["Rm"]]` is an $N_y x N_y$ matrix.

`ssm[["betaO"]]` is an optional $N_y x N_o$ matrix.

`ssm[["betaS"]]` is an optional $N_{\beta} x N_s$ matrix.

To force exclusion of exogenous data, set the relevant matrices to have all 0 values. This is the default. 

To use an unweighted likelihood, set the weights to be all 1's. This is the default.

`smooth` is a boolean indicating whether to run the backwards smoother or not.

The output gives a list with values `lnl` (the log likelihood), `y_tl` (the predicted values of `y` given information up to time $t-1$), `y_tt` (the predicted values of `y` using all the available information), `B_tl` (the predicted values of the unobserved components up to time $t-1$), `B_tt` (the predicted values of the unobserved components given all the available information), `P_tl` (the unobserved components covariance matrix up to time $t-1$), `P_tt` (the unobserved components covariance matrix using all the available information), `F_t` (variance of the prediction error of `y` up to time $t-1$), `N_t` (the prediction error of `y` up to time $t-1$), and `K_t` (the Kalman gain). 

## Example: Nelson-Siegel Dynamic Factor Yield Curve Model

```{r}
library(kalmanfilter)
library(maxLik)
library(ggplot2)
library(data.table)
library(gridExtra)
data(treasuries)
tau = unique(treasuries$maturity)

#State space model for the Nelson-Siegel dynamic factor yield curve
yc_ssm = function(par, tau){
  #Set factor names
  tau = unique(tau)
  factors = c("l", "s", "c")
  
  #Loading on the slope factor
  Ls = function(par, tau){
    return(-(1 - exp(-tau*par["lambda"]))/(tau*par["lambda"]))
  }
  
  #Loading on the curvature factor
  Lc = function(par, tau){
    return(-Ls(par["lambda"], tau) - exp(-tau*par["lambda"]))
  }
  
  ########## Define the state space model
  #Transition matrix
  Fm = matrix(par[grepl("phi_", names(par))], nrow = 3, ncol = 3,
              dimnames = list(paste0(factors, "_t0"),
                              paste0(factors, "_t1")))
  Fm[is.na(Fm)] = 0
  
  #Transition intercept matrix
  Dm = matrix(par[grepl("D_", names(par))], ncol = 1,
              dimnames = list(rownames(Fm), NULL))
  
  #Transition error covariance matrix
  Qm = matrix(par[unlist(lapply(factors, function(x){paste0("sig_", x, factors)}))], nrow = 3, ncol = 3)
  Qm[lower.tri(Qm)] = Qm[upper.tri(Qm)]
  dimnames(Qm) = list(rownames(Fm), rownames(Fm))
  
  #Observation equation matrix
  n = length(tau)
  Hm = matrix(NA, nrow = n, ncol = 3, 
              dimnames = list(as.character(tau), rownames(Fm)))
  tau = as.numeric(tau)
  if("l" %in% factors){
    Hm[, "l_t0"] = 1
  }
  if("s" %in% factors){
    Hm[, "s_t0"] = Ls(par, tau)
  }
  if("c" %in% factors){
    Hm[, "c_t0"] = Lc(par, tau)
  }

  #Observation error variance covariance matrix
  Rm = diag(sum(grepl("sig_\\d+", names(par))))*par[grepl("sig_\\d+", names(par))]^2
  dimnames(Rm) = list(rownames(Hm), rownames(Hm))
  
  #Observation intercept matrix
  Am = matrix(0, nrow = nrow(Hm), ncol = 1, 
              dimnames = list(rownames(Hm), NULL))
  
  #Initial guess for the unobserved vector
  B0 = matrix(par[names(par) %in% c("level", "slope", "curvature")], ncol = 1, nrow = nrow(Fm), 
              dimnames = list(rownames(Fm), NULL))
  
  #Initial guess for the variance of the unobserved vector
  P0 = diag(par["P0"]^2, nrow = nrow(Qm), ncol = ncol(Qm))
  
  return(list(Fm = Fm, Dm = Dm, Qm = Qm, 
              Hm = Hm, Am = Am, Rm = Rm, 
              B0 = B0, P0 = P0))
}

#Set the initial values
init = c(level = 5.9030, slope = -0.7090, curvature = 0.8690, lambda = 0.0423, 
         D_l = 0.1234, D_s = -0.2285, D_c = 0.2020, 
         phi_ll = 0.9720, phi_ls = 0.1009, phi_lc = -0.1226, 
         phi_sl = -0.0209, phi_ss = 0.8189, phi_sc = 0.0192, 
         phi_cl = -0.0061, phi_cs = -0.1446, phi_cc = 0.8808, 
         sig_ll = 0.1017, 
         sig_sl = 0.0937, sig_ss = 0.2267,
         sig_cl = 0.0303, sig_cs = 0.0351, sig_cc = 0.7964, 
         sig_1 = 0.0934, sig_3 = 0.0001, sig_6 = 0.1206, sig_12 = 0.1525, 
         sig_24 = 0.1328, sig_36 = 0.0855, sig_60 = 0.0001, sig_84 = 0.0397, 
         sig_120 = 0.0595, sig_240 = 0.1438, sig_360 = 0.1450, 
         P0 = 0.0001)

#Set up the constraints: lambda and all standard deviation parameters must be positive
ineqA = matrix(0, nrow = 15, ncol = length(init), dimnames = list(NULL, names(init)))
diag(ineqA[, c("lambda", "sig_ll", "sig_ss", "sig_cc", paste0("sig_", tau))]) = 1
ineqB = matrix(0, nrow = nrow(ineqA), ncol = 1)

#Convert to an NxT matrix
yt = dcast(treasuries, "date ~ maturity", value.var = "value")
yt = t(yt[, 2:ncol(yt)])

#Set the objective function
objective = function(par){
  ssm = yc_ssm(par, tau)
  return(kalman_filter(ssm, yt,)$lnl)
}

#Solve the model
solve = maxLik(logLik = objective, start = init, method = "BFGS",
               finalHessian = FALSE, hess = NULL,
               control = list(printLevel = 2, iterlim = 10000), 
               constraints = list(ineqA = ineqA, ineqB = ineqB))

#Get the estimated state space model
ssm = yc_ssm(solve$estimate, tau)

#Filter the data with the model
filter = kalman_filter(ssm, yt, smooth = TRUE)

#Get the estimated unobserved factors
B_tt = data.table(t(filter[["B_tt"]]))[, "date" := unique(treasuries$date)]
colnames(B_tt)[1:3] = c("level", "slope", "curvature")

#Get the estimated yields
y_tt = data.table(t(filter[["y_tt"]]))[, "date" := unique(treasuries$date)]
colnames(y_tt)[1:(ncol(y_tt) - 1)] = tau

#Plot the data
g1 = ggplot(treasuries, id.vars = "date") + 
  geom_line(aes(x = date, y = value, group = factor(maturity), color = factor(maturity))) + 
  theme_minimal() + theme(legend.position = "bottom") +
  labs(title = "Actual Treasury Yields", x = "", y = "Value") + 
  guides(color = guide_legend(title = NULL))
g2 = ggplot(melt(y_tt, id.vars = "date")) + 
  geom_line(aes(x = date, y = value, group = variable, color = variable)) + 
  theme_minimal() + theme(legend.position = "bottom") +
  labs(title = "Estimated Treasury Yields", x = "", y = "Value") + 
  guides(color = guide_legend(title = NULL))
g3 = ggplot(melt(B_tt, id.vars = "date")) + 
  geom_line(aes(x = date, y = value, group = variable, color = variable)) + 
  theme_minimal() + theme(legend.position = "bottom") +
  labs(title = "Estimated Factors", x = "", y = "Value") + 
  guides(color = guide_legend(title = NULL))
grid.arrange(g1, g2, g3, layout_matrix = rbind(c(1, 2), c(3, 3)))
```

### With Time-Varying Parameters

```{r}
#Set the objective function
objective = function(par){
  ssm = yc_ssm(par, tau)
  for(x in names(ssm)){
    if(!x %in% c("B0", "P0")){
      ssm[[x]] = array(ssm[[x]], dim = c(dim(ssm[[x]]), ncol(yt)))
    }
  }
  return(kalman_filter(ssm, yt)$lnl)
}

#Solve the model
solve = maxLik(logLik = objective, start = init, method = "BFGS",
               finalHessian = FALSE, hess = NULL,
               control = list(printLevel = 2, iterlim = 10000), 
               constraints = list(ineqA = ineqA, ineqB = ineqB))
```

## Stock and Watson Dynamic Common Factor Model

```{r}
library(kalmanfilter)
library(data.table)
library(maxLik)
library(ggplot2)
library(gridExtra)
data(sw_dcf)
data = sw_dcf[, colnames(sw_dcf) != "dcoinc", with = F]
vars = colnames(data)[colnames(data) != "date"]

#State space model for the Stock and Watson Dynamic Common Factor model
dcf_ssm = function(par, yt){
  #Get the parameters
  vars = dimnames(yt)[which(unlist(lapply(dimnames(yt), function(x){!is.null(x)})))][[1]]
  phi = par[grepl("phi", names(par))]
  names(phi) = gsub("phi", "", names(phi))
  gamma = par[grepl("gamma_", names(par))]
  names(gamma) = gsub("gamma_", "", names(gamma))
  psi = par[grepl("psi_", names(par))]
  names(psi) = gsub("psi_", "", names(psi))
  sig = par[grepl("sigma_", names(par))]
  names(sig) = gsub("sigma_", "", names(sig))
  
  #Build the transition matrix 
  phi_dim = max(c(length(phi)), length(unique(sapply(strsplit(names(gamma), "\\."), function(x){x[2]}))))
  psi_dim = sapply(unique(sapply(strsplit(names(psi), "\\."), function(x){x[1]})), function(x){
    max(as.numeric(sapply(strsplit(names(psi)[grepl(paste0("^", x), names(psi))], "\\."), function(x){x[2]})))
  })
  Fm = matrix(0, nrow = phi_dim + length(psi), ncol = phi_dim + length(psi), 
              dimnames = list(
                c(paste0("ct.", 0:(phi_dim - 1)), 
                  unlist(lapply(names(psi_dim), function(x){paste0("e_", x, ".", 0:(psi_dim[x] - 1))}))),
                c(paste0("ct.", 1:phi_dim), 
                  unlist(lapply(names(psi_dim), function(x){paste0("e_", x, ".", 1:psi_dim[x])})))
              ))
  Fm["ct.0", paste0("ct.", names(phi))] = phi
  for(i in 1:length(vars)){
    Fm[paste0("e_", i, ".0"), 
       paste0("e_", names(psi)[grepl(paste0("^", i), names(psi))])] = psi[grepl(paste0("^", i), names(psi))]
  }
  diag(Fm[intersect(rownames(Fm), colnames(Fm)), intersect(rownames(Fm), colnames(Fm))]) = 1
  
  #Build the observation matrix
  Hm = matrix(0, nrow = nrow(yt), ncol = nrow(Fm), dimnames = list(rownames(yt), rownames(Fm)))
  for(i in 1:length(vars)){
    Hm[i, paste0("ct.", sapply(strsplit(names(gamma)[grepl(paste0("^", i), names(gamma))], "\\."), function(x){x[2]}))] = 
      gamma[grepl(paste0("^", i), names(gamma))]
  }
  diag(Hm[, paste0("e_", 1:length(vars), ".0")]) = 1
  
  #Build the state covariance matrix
  #Set the dynamic common factor standard deviation to 1
  Qm = matrix(0, ncol = ncol(Fm), nrow = nrow(Fm), dimnames = list(rownames(Fm), rownames(Fm)))
  Qm["ct.0", "ct.0"] = 1
  for(i in 1:length(vars)){
    Qm[paste0("e_", i, ".0"), paste0("e_", i, ".0")] = sig[names(sig) == i]^2
  }
  
  #Build the observation equation covariance matrix
  Rm = matrix(0, ncol = nrow(Hm), nrow = nrow(Hm), dimnames = list(vars, vars))
  
  #Transition equation intercept matrix
  Dm = matrix(0, nrow = nrow(Fm), ncol = 1, dimnames = list(rownames(Fm), NULL))
  
  #Observation equation intercept matrix
  Am = matrix(0, nrow = nrow(Hm), ncol = 1)
  
  #Initialize the filter for each state
  B0 = matrix(0, nrow(Fm), 1)
  P0 = diag(nrow(Fm))
  dimnames(B0) = list(rownames(Fm), NULL)
  dimnames(P0) = list(rownames(Fm), rownames(Fm))
  
  B0 = solve(diag(ncol(Fm)) - Fm) %*% Dm
  VecP_ll = solve(diag(prod(dim(Fm))) - kronecker(Fm, Fm)) %*% matrix(as.vector(Qm), ncol = 1)
  P0 = matrix(VecP_ll[, 1], ncol = ncol(Fm))

  return(list(B0 = B0, P0 = P0, Am = Am, Dm = Dm, Hm = Hm, Fm = Fm, Qm = Qm, Rm = Rm))
}

#Log the data
data.log = copy(data)
data.log[, c(vars) := lapply(.SD, log), .SDcols = c(vars)]

#Difference the data
data.logd = copy(data.log)
data.logd[, c(vars) := lapply(.SD, function(x){
  x - shift(x, type = "lag", n = 1)
}), .SDcols = c(vars)]

#Standardize the data
data.logds = copy(data.logd)
data.logds[, c(vars) := lapply(.SD, scale, scale = FALSE), .SDcols = c(vars)]

#Transpose the data
yt = t(data.logds[, c(vars), with = FALSE])

#Set the initial values
init = c(phi1 = 0.8588, phi2 = -0.1526, 
         psi_1.1 = -0.1079, psi_1.2 = -0.1415,
         psi_2.1 = -0.3166, psi_2.2 = -0.0756,
         psi_3.1 = -0.3994, psi_3.2 = -0.2028,
         psi_4.1 = -0.0370, psi_4.2 = 0.3646,
         gamma_1.0 = 0.0064, gamma_1.1 = -0.0020,
         gamma_2.0 = 0.0018, 
         gamma_3.0 = 0.0059, gamma_3.1 = -0.0027,
         gamma_4.0 = 0.0014, gamma_4.1 = -0.0004, gamma_4.2 = 0.0001, gamma_4.3 = 0.0004,
         sigma_1 = 0.0049, sigma_2 = 0.0056, sigma_3 = 0.0077, sigma_4 = 0.0013)

#Set the constraints
ineqA = matrix(0, nrow = 14, 
               ncol = length(init), dimnames = list(NULL, names(init)))
#Stationarity constraints
ineqA[c(1, 2), c("phi1", "phi2")] = rbind(c(1, 1), c(-1, -1))
ineqA[c(3, 4), grepl("psi_1", colnames(ineqA))] = rbind(c(1, 1), c(-1, -1))
ineqA[c(5, 6), grepl("psi_2", colnames(ineqA))] = rbind(c(1, 1), c(-1, -1))
ineqA[c(7, 8), grepl("psi_3", colnames(ineqA))] = rbind(c(1, 1), c(-1, -1))
ineqA[c(9, 10), grepl("psi_4", colnames(ineqA))] = rbind(c(1, 1), c(-1, -1))
#Non-negativity constraints
diag(ineqA[c(11, 12, 13, 14), grepl("sigma_", colnames(ineqA))]) = 1
ineqB = matrix(c(rep(1, 10), 
                 rep(0, 4)), nrow = nrow(ineqA), ncol = 1)

#Define the objective function
objective = function(par, yt){
  ssm = dcf_ssm(par, yt)
  return(kalman_filter(ssm, yt, smooth = FALSE)$lnl)
}

#Solve the model
solve = maxLik(logLik = objective, start = init, method = "BFGS", 
               finalHessian = FALSE, hess = NULL, 
               control = list(printLevel = 2, iterlim = 10000), 
               constraints = list(ineqA = ineqA, ineqB = ineqB), 
               yt = yt)

#Get the estimated state space model
ssm = dcf_ssm(solve$estimate, yt)

#Get the column means and standard deviations
M = matrix(unlist(data.logd[, lapply(.SD, mean, na.rm = TRUE), .SDcols = c(vars)]), 
               ncol = 1, dimnames = list(vars, NULL))

#Get the coefficient matrices
Hm = ssm[["Hm"]]
Fm = ssm[["Fm"]]

#Final K_t is approximation to steady state K
filter = kalman_filter(ssm, yt, smooth = T)
K = filter$K_t[,, dim(filter$K_t)[3]]
W = solve(diag(nrow(K)) - (diag(nrow(K)) - K %*% Hm) %*% Fm) %*% K
d = (W %*% M)[1, 1]

#Get the intercept terms
gamma = Hm[, grepl("ct", colnames(Hm))]
D = M - gamma %*% matrix(rep(d, ncol(gamma)))

#Initialize first element of the dynamic common factor
Y1 = t(data.log[, c(vars), with = F][1, ])
initC = function(par){
  return(sum((Y1 - D - gamma %*% par)^2))
}
C10 = optim(par = Y1, fn = initC, method = "BFGS", control = list(trace = FALSE))$par[1]
Ctt = rep(C10, ncol(yt))

#Build the rest of the level of the dynamic common factor
ctt = filter$B_tt[which(rownames(Fm) == "ct.0"), ]
for(j in 2:length(Ctt)){
  Ctt[j] = ctt[j] + Ctt[j - 1] + c(d)
}
Ctt = data.table(date = data$date, dcf = Ctt, d.dcf = ctt)

#Plot the outputs
g1 = ggplot(melt(data.log, id.vars = "date")[, "value" := scale(value), by = "variable"]) + 
  ggtitle("Data", subtitle = "Log Levels (Rescaled)") + 
  scale_y_continuous(name = "Value") + 
  scale_x_date(name = "") + 
  geom_line(aes(x = date, y = value, group = variable, color = variable)) + 
  theme_minimal() + theme(legend.position = "bottom") + guides(color = guide_legend(title = NULL))

g2 = ggplot( melt(data.logds, id.vars = "date")) + 
  ggtitle("Data", subtitle = "Log Differenced & Standardized") + 
  scale_y_continuous(name = "Value") + 
  scale_x_date(name = "") + 
  geom_hline(yintercept = 0, color = "black") + 
  geom_line(aes(x = date, y = value, group = variable, color = variable)) + 
  theme_minimal() + theme(legend.position = "bottom") + guides(color = guide_legend(title = NULL))

g3 = ggplot(melt(Ctt, id.vars = "date")[variable == "dcf", ]) +  
  ggtitle("Dynamic Common Factor", subtitle = "Levels") + 
  scale_x_date(name = "") +
  geom_hline(yintercept = 0, color = "grey") + 
  geom_line(aes(x = date, y = value, group = variable, color = variable)) + 
  theme_minimal() + theme(legend.position = "bottom") + scale_color_manual(values = "black") + 
  guides(color = guide_legend(title = NULL), fill = guide_legend(title = NULL)) + 
  scale_y_continuous(name = "Value", limits = range(Ctt$dcf, na.rm = TRUE))

g4 = ggplot(melt(Ctt, id.vars = "date")[variable == "d.dcf", ]) +  
  ggtitle("Dynamic Common Factor", subtitle = "Differenced") + 
  scale_x_date(name = "") +
  geom_hline(yintercept = 0, color = "grey") + 
  geom_line(aes(x = date, y = value, group = variable, color = variable)) + 
  theme_minimal() + theme(legend.position = "bottom") + scale_color_manual(values = "black") +
  guides(color = guide_legend(title = NULL), fill = guide_legend(title = NULL)) + 
  scale_y_continuous(name = "Value", limits = range(Ctt$d.dcf, na.rm = TRUE))

grid.arrange(g1, g2, g3, g4, layout_matrix = matrix(c(1, 3, 2, 4), nrow = 2))
```
